---
title: "RMS Titanic"
author: "Elisa Camacho"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
glimpse(df_titanic)
## TASK: Perform a `glimpse` of df_titanic
```

**Observations**:

-   Class
-   Age
-   Sex
-   Survived
-   n  (Count of passengers in each category)

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Write your observations here
The total count from the dataset is 2201 passengers and crew.

-   Are there any differences?
According to Wikipedia, there were approximately 2,224 people aboard.
The difference is about 23 people (1% difference).

-   If yes, what might account for those differences?
The dataset might have excluded some crew members or passengers with incomplete records. 
Counting methods may be different (some sources might count babies differently)
    -   ...

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
survivors <- df_titanic %>% filter(Survived == "Yes")
survivors %>% 
  ggplot(aes(fill = Sex, x = Class, y = n)) +
  geom_col(position = "Stack")
```

**Observations**:

Female survivors outnumbered male survivors in all passenger classes except the crew members

Crew members had a significant number of survivors, particularly the male crew

1st class had more survivors than 2nd class and 3rd class

For 3rd class their is an even split between male and female survivors. 

"women and children first"

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge") +
  labs(title = "Proportion of Survivors by Class and Sex",
       x = "Passenger Class",
       y = "Proportion Survived")
```

**Observations**:

-   Write your observations here.

Females had much higher survival rates than males across all classes

1st class passengers had the highest survival rates, same thing with 2nd class, then 3rd class

-   Is there anything *fishy* going on in your plot?
this graph 1st and 2nd class have the same survivors, while that is not true.

while 3rd class had the most survivors in absolute numbers, they actually had the lowest survival rates when considering their total numbers

Crew members, especially males, had very low survival rates despite their high absolute numbers in q3

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_grid(~ Age) +
  labs(title = "Proportion of Survivors by Class, Sex, and Age",
       x = "Passenger Class",
       y = "Proportion Survived")
```

**Observations**:

-   (Write your observations here.)
Children generally had higher survival rates than adults across all classes

The "women and children first" policy is clearly visible in the data

Female children in 1st and 2nd class had nearly 100% survival rates

Male children also had relatively high survival rates compared to male adults, most likely because their children. 

-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.

the high raw counts in 3rd class were likely due to more children being in that class

Crew members had very few children (as expected), so their survival patterns are dominated by adult rates

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
