---
title: "Michelson Speed-of-light Measurements"
author: "Elisa Camacho"
date: 2020-
output: 
  github_document:
    toc: true
prerequisites:
  - e-data02-derive
---

*Purpose*: When studying physical problems, there is an important distinction between *error* and *uncertainty*. The primary purpose of this challenge is to dip our toes into these factors by analyzing a real dataset.

*Reading*: [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (Optional)

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to measure the speed of light. His approach was a development upon the method of Foucault[3], and resulted in a new estimate of $v_0 = 299944 \pm 51$ kilometers per second (in a vacuum). This is very close to the modern *exact* value of `r LIGHTSPEED_VACUUM`. In this challenge, you will analyze Michelson's original data, and explore some of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code chunk below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the value `LIGHTSPEED_VACUUM`) because the meter is actually [*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition) in terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse()
```

*Data dictionary*:

- `Date`: Date of measurement
- `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair, 1 = poor
- `Temp`: Ambient temperature (Fahrenheit)
- `Velocity`: Measured speed of light (km / s)

### __q1__ Re-create the following table (from Michelson (1880), pg. 139) using `df_michelson` and `dplyr`. Note that your values *will not* match those of Michelson *exactly*; why might this be?

| Distinctness | n  | MeanVelocity |
|--------------|----|----------|
|            3 | 46 |   299860 |
|            2 | 39 |   299860 |
|            1 | 15 |   299810 |

```{r q1-task}
## TODO: Compute summaries
## Distinctness of measured images: 3 = good, 2 = fair, 1 = poor
df_q1 <- df_michelson %>%
  group_by(Distinctness) %>%
  summarise(
    n = n(),
    MeanVelocity = mean(Velocity, na.rm = TRUE) 
  ) %>%
  arrange(desc(Distinctness))

df_q1 %>%
  knitr::kable(digits = 0)

```

**Observations**:
- Write your observations here!
The mean velocities are nearly identical for distinctness levels 3 (good) and 2 (fair), but slightly lower for level 1 (poor)

The number of measurements decreases with quality (46 → 39 → 15)

The small discrepancies (e.g., 299860 vs Michelson's 299861) are likely due to:
Rounding differences (Michelson may have rounded intermediate calculations differently)

Possible transcription differences in the historical record

The exact dataset Michelson used for this table may have been slightly different

### __q2__ Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to `Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <- df_michelson %>% 
  mutate(VelocityVacuum = Velocity + 92)

df_q2
```

As part of his study, Michelson assessed the various potential sources of error,
and provided his best-guess for the error in his speed-of-light estimate. These
values are provided in `LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently, Michelson
believed the true value of the speed-of-light probably lay between
`LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and `LIGHTSPEED_MICHELSON +
LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

- **Error** is the difference between a true value and an estimate of that
  value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
- **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does not know
the error. The best they can do is quantify their degree of uncertainty. We will
learn some means of quantifying uncertainty in this class, but for many real
problems uncertainty includes some amount of human judgment.[2]

### __q3__ Compare Michelson's speed of light estimate against the modern speed of light value. Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
## Your code here!
LIGHTSPEED_VACUUM <- 299792.458  
LIGHTSPEED_MICHELSON <- 299944.00  
LIGHTSPEED_PM <- 51  

true_error <- abs(LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON)
uncertainty_range <- c(LIGHTSPEED_MICHELSON - LIGHTSPEED_PM, LIGHTSPEED_MICHELSON + LIGHTSPEED_PM)
is_within_range <- LIGHTSPEED_VACUUM >= uncertainty_range[1] & LIGHTSPEED_VACUUM <= uncertainty_range[2]

# Print results
print(paste("True Error (km/s):", true_error))
print(paste("Michelson's Uncertainty Range (km/s):", paste(uncertainty_range, collapse = " to ")))
print(paste("Is True Value Within Uncertainty Range?:", is_within_range))
```

**Observations**:
- Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?
  - Michelson's estimate: 299,944.00 ± 51 km/s
  - True value: 299,792.458 km/s

- Make a quantitative comparison between Michelson's uncertainty and his error.
  - True error: 299,944.00 - 299,792.458 = 151.542 km/s
  - Michelson's uncertainty range: 299,893 to 299,995 km/s
Michelson's uncertainty (±51 km/s) was smaller than the true error (151.542 km/s)

The true value (299,792.458 km/s) falls outside Michelson's uncertainty range

Michelson underestimated his measurement error by about 3× (151.542/51 ≈ 3)

The following plot shows all of Michelson's data as a [control chart](https://en.wikipedia.org/wiki/Control_chart); this sort of plot is common in manufacturing, where it is used to help determine if a manufacturing process is under [statistical control](https://en.wikipedia.org/wiki/Statistical_process_control). Each dot is one of Michelson's measurements, and the grey line connects the mean taken for each day. The same plot also shows simulated data using a probability model. We'll get into statistics later in the course; for now, let's focus on understanding what real and simulated data tend to look like.

### __q4__ Inspect the following plot with the `Real` Michelson data and `Simulated` data from a probability model. Document the similarities and differences between the data under *observe* below.

```{r q4-cf-real-simulated}
## Note: No need to edit this chunk!
## Calibrate simulated data
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%

  ggplot(aes(Date, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +

  geom_line(
    data = . %>%
      group_by(Date, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Date of Measurement (1879)",
    y = "Velocity (in Vacuum)"
  )
```

**Observations**:
Similarities:
Both datasets show variation around Michelson's mean estimate (dotted line)

Most points in both datasets fall within Michelson's uncertainty bounds (dashed lines)

The general shape of fluctuations is similar between real and simulated data

Differences:
Real data has more extreme values (point near 300,100 km/s in June)

Simulated data shows tighter clustering around the mean

Real data shows clear time-based patterns

Real data shows quality changes over time

Real data shows consistent overestimation relative to true value

Simulated data centers on Michelson's (incorrect) mean

### __q5__ You have access to a few other variables. Construct a **at least three** visualizations of `VelocityVacuum` against these other factors. Are there other patterns in the data that might help explain the difference between Michelson's estimate and `LIGHTSPEED_VACUUM`?
```{r q5}
glimpse(df_q2)

#Graph 1
ggplot(df_q2, aes(x = Temp, y = VelocityVacuum)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = LIGHTSPEED_VACUUM, 
             color = "red", 
             linetype = "dashed") +
  labs(title = "Speed vs. Temperature",
       x = "Temperature (°F)",
       y = "Measured Speed (km/s)") +
  theme_minimal()

#Graph 2
ggplot(df_q2, aes(x = Distinctness, y = VelocityVacuum)) +
  geom_boxplot(fill = "lightblue") +
  geom_hline(yintercept = LIGHTSPEED_VACUUM, 
             color = "red", 
             linetype = "dashed") +
  labs(title = "Measurement Quality Comparison",
       x = "Distinctness (3=Best)",
       y = "Measured Speed (km/s)") +
  theme_minimal()

#Graph 3
df_q2 %>% 
  group_by(Temp) %>% 
  ggplot() + 
  aes(Temp, VelocityVacuum) +
  geom_point() +
  ggtitle("Velocity in a Vacuum vs. Temperature")

#Graph 4
df_q2 %>% 
  ggplot(aes(x = VelocityVacuum, color = Distinctness)) +
  geom_bar() +
  ggtitle("Velocity in a Vacuum Count/Distinctness")

```

**Observations**:

- Make sure to record observations on your graphs!
For Graph 1 Speed vs. Temperature
- Points show a slight downward trend as temperature increases
- All measurements are above the true speed of light (red line)
- No extreme outliers in the temperature range

For Graph 2 Measurement Quality Comparison

- All quality groups show similar median values
- True speed (red line) falls below all measurements

For Graph 3 Velocity in a Vacuum vs. Temperature. 

- The peak velocity measurement occurs in the middle of the temperature values - However, the lowest velocity measurement occurs at the lower end of the temperature range

For Graph 4 Velocity in a Vacuum Count/Distinctness 
- Higher velocity measurements tend to be 2 and 3 level distinctness. 
- 1 is the least common distinctness value

## Bibliography

- [1] Michelson, [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (1880)
- [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf) (1986)
- [3] BYU video about a [Fizeau-Foucault apparatus](https://www.youtube.com/watch?v=Ik5ORaaeaME), similar to what Michelson used.
