---
title: "Gapminder"
author: "Elisa Camacho"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: Learning to do EDA well takes practice! In this challenge
you'll further practice EDA by first completing a guided exploration,
then by conducting your own investigation. This challenge will also give
you a chance to use the wide variety of visual tools we've been
learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and
supporting files (`report_files/` folder) when you are done! Then submit
a link to Canvas. **Your Challenge submission is not complete without
all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is
an independent organization that seeks to educate people about the state
of the world. They seek to counteract the worldview constructed by a
hype-driven media cycle, and promote a "fact-based worldview" by
focusing on data. The dataset we'll study in this challenge is from
Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to
the high-level process we're going through---after this guided round
you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
```

**Observations**:

Write all variable names here

-   country: Nation name

-   continent: Continent

-   year: Year of observation

-   lifeExp: Life expectancy at birth

-   pop: Total population

-   gdpPercap: GDP per capita (US dollars, inflation-adjusted)

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble.
(Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
#year_max <- NA_real_
#year_min <- NA_real_
# Find the most recent and least recent years
year_max <- gapminder %>% pull(year)  %>% max()

year_min <- gapminder %>% pull(year)  %>% min()


# results
year_max
year_min
```
Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document
your challenges and attempt to produce the most informative visual you
can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>% 
  filter(year == year_min) %>% 
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() + 
  scale_y_log10() +
  labs(title = "GDP per Capita by Continent (1952)",
       y = "GDP per capita (log scale)")
```

**Observations**: There are a few outliers in Asia and Europe, where
some countries have significantly higher GDP per capita than the rest.
Oceania and Europe have higher median GDP per capita compared to other
continents. Africa has the lowest median GDP per capita. The spread of
GDP per capita is widest in Asia, indicating significant economic
diversity. Europe has a relatively narrow spread, suggesting more
economic homogeneity. Difficulties & Approaches: Some countries have
extremely high GDP per capita, which can skew the plot.

Revised
The log scale reveals more differences between continents. 
Africa's GDP distribution becomes more visible (median about $1,000)
The extreme outlier (Kuwait) is even more apparent
Europe's interquartile range appears narrower on log scale

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder %>% 
  filter(year == year_min) %>% 
  filter(continent == "Americas" | continent == "Asia") %>% 
  arrange(desc(gdpPercap))
```

**Observations**:

-   Identify the outlier countries from q2
    -   Kuwait (Asia extreme outlier), United States(Americas), and Canada (Americas)

*Hint*: For the next task, it's helpful to know a ggplot trick we'll
learn in an upcoming exercise: You can use the `data` argument inside
any `geom_*` to modify the data that will be plotted *by that geom
only*. For instance, you can use this trick to filter a set of points to
label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple
variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>% 
  filter(year %in% c(year_min, year_max)) %>% 
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  scale_y_log10() +
  facet_grid(~year) +
  geom_point(
    data = . %>% filter(country %in% c("Kuwait", "United States", "Canada")),
    color = "red",
    size = 2
  ) +
  labs(title = "GDP Comparison: 1952 vs 2007",
       y = "GDP per capita (log scale)")

```

**Observations**: The comparison between 1952 and 2007 shows significant
economic growth across all continents.

The outliers in 1952 (Kuwait) highlight countries that were already
economically advanced at the time.

The rise of new outliers in 2007 reflects the new economic growth.

Revised Observations 
- Using facet_grid provides better vertical space for comparison
- All continents show upward shift in GDP distributions
- Africa's growth appears modest compared to others
- Asia shows dramatic expansion in upper range
- The Americas' distribution became more compressed
- Europe maintained relatively tight distribution
- Outliers in 2007 are less extreme relative to peers

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP
per capita at two time points. You can continue looking at outliers,
consider different years, repeat the exercise with `lifeExp`, consider
the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>%
  ggplot(aes(x = year, y = lifeExp, color = continent)) +
  geom_smooth(se = FALSE) +
  labs(title = "Life Expectancy Trends by Continent")
```

-   Observations
-   Life expectancy has increased across all continents from 1952 to
    2007. 
-   Europe and Oceania have the highest life expectancy, while Africa has
the lowest.

Questions:

How has the life expectancy gap between continents changed over time?

Which continent showed the most consistent growth pattern?

```{r q5-task2}
## TASK: Your second graph
# Plot GDP per capita vs life expectancy
gapminder %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  facet_wrap(~continent) +
  labs(title = "GDP vs Life Expectancy by Continent")
```

-   (Your notes and observations here)

The relationship shown here is that There is a positive correlation
between GDP per capita and life expectancy.

Countries with higher GDP per capita tend to have higher life
expectancy.

Some countries (Africa) have low GDP per capita but relatively high life
expectancy.

Some countries (Asia) have high GDP per capita but lower-than-expected
life expectancy.

Questions:

Does the range in life expectancy in Asia have a correlation with having a large population? 

```{r q5-task3}
## TASK: Your third graph
# Plot population growth by continent over time
gapminder %>%
  group_by(continent, year) %>%
  summarise(total_pop = sum(pop)) %>%  # Calculate total population per continent
  ggplot(aes(x = year, y = total_pop, color = continent)) +
  geom_line() +  # Add line plot
  labs(
    title = "Population Growth by Continent Over Time",
    x = "Year",
    y = "Total Population",
    color = "Continent"
  ) +
  theme_minimal()
```

-   (Your notes and observations here)
-   A trend I see is that Asia has the highest population and the
    fastest growth rate.
-   Africa shows rapid population growth starting in the 1980s.
-   Europe has relatively stable population growth.

Questions:
Which continents show accelerating vs decelerating growth?

How does population growth rate correlate with economic indicators?